---
title: "EPC_2026_Data_Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EPC_2026_Data_Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the packages that we will be using for this process:

```{r setup}
#install_from github here
library(dplyr)
library(readr)
library(stringr)
library(dbplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(forcats)
```

Create a base population of 5000 individuals, taking into account age, fertility and homeownership.
For this workshop, we'll also only show one country, Spain.

For this example we only use publicly available, open access data. This is to show you one path of building an initial population, even when you may not have all the data you would want. In a lot of cases you probably have access to microdata and can build your initial population from that.

But for this example, all the data can also be downloaded here:

https://ec.europa.eu/eurostat/databrowser/view/demo_pjan/default/table?lang=en&category=demo.demo_pop

https://ec.europa.eu/eurostat/databrowser/view/ilc_lvho02__custom_21562534/default/table

All cleaned data are available in the package, but the cleaning can be found here:

```{r load and prepare raw data, eval=FALSE, echo=FALSE}
#invisible(gc())
#rm(list = ls())
# ============================================================================
# 1. LOAD AND PREPARE RAW DATA
# ============================================================================

# Load population data (by age and sex)
# path <- system.file("extdata", "EPC_2026", "demo_pjan__custom_21562150_linear_2_0.csv", package = "MicSim")
# df_pjan_raw <- read_csv(path,col_types = cols(.default = col_character()))
#save(df_pjan_raw, file = "df_pjan_raw.rda")

# Load housing data (tenure status by household composition)
# path <- system.file("extdata", "EPC_2026", "ilc_lvho02__custom_21562534_linear_2_0.csv", package = "MicSim")
# df_housing_raw <- read_csv(path,col_types = cols(.default = col_character()))
#save(df_housing_raw, file = "df_housing_raw.rda")
# ============================================================================
# Preparation of Raw Data
# ============================================================================
data(df_pjan_raw)
data(df_housing_raw)
unique(df_pjan_raw$age)
unique(df_pjan_raw$sex)
head(df_pjan_raw)
data(pop_structure)

# For now keep only Spain and clean (most complete data )
# pop_structure <- df_pjan_raw %>%
#   filter(
#     `Geopolitical entity (reporting)`  == "Spain",
#     !is.na(`OBS_VALUE`),
#     age != "TOTAL",
#     age != "UNK"
#   ) %>%
#   mutate(
#     population = as.numeric(`OBS_VALUE`),
#     year = as.integer(`TIME_PERIOD`)
#   ) %>%
#   filter(!is.na(age), !is.na(sex), !is.na(population)) %>%
#   select(year, age, sex, population)

#save(pop_structure, file = "pop_structure.rda")

# Extract Spain housing data: home ownership rates

head(df_housing_raw$OBS_VALUE[df_housing_raw$`Household composition` == "Total"])

housing_es <- df_housing_raw %>%
  filter(
    `Geopolitical entity (reporting)` == "Spain",
    hhcomp == "TOTAL",
    !is.na(`OBS_VALUE`),
  ) %>%
  mutate(
    obs = `OBS_VALUE`,
    year = as.integer(`TIME_PERIOD`)
  ) %>%
  select(year, tenure, obs)

#save(housing_es, file = "ilc_lvho02_sp_cl.rda")
```

If the first step is done, load in cleaned data (also available in the package):

```{r load .rda files}
invisible(gc())
rm(list = ls())

data(pjan_clean)
data(ilc_lvho02_sp_cl)

set.seed(321342)
```


```{r clean age}
#### Clean Age ####
data(pop_structure)
# unique(pop_structure$age)

# pop_es <- pop_structure %>% 
#   filter(!age %in% c("Y_LT1", "Y_OPEN")) %>% 
#   mutate(age = as.numeric(str_remove(age, "^Y")))
# 
# save(pop_es, file = "pop_es.rda")
data(pop_es)
```

First we want averages of the population per age and sex.

```{r filter data}
# Age distribution
pop_avg <- pop_es %>%
  filter(sex %in% c("F", "M")) %>%
  group_by(age, sex) %>%
  summarise(
    population = mean(population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(sex = tolower(sex)) #some more sneaky cleaning

print(pop_avg %>% 
        group_by(sex) %>% 
        summarise(avg_pop = sum(population, na.rm = TRUE), .groups = "drop"))

# Create sampling weights
pop_weights <- pop_avg %>%
  mutate(
    total_pop = sum(population, na.rm = TRUE),
    weight = population / total_pop
  ) %>%
  select(age, sex, weight)
```

Next we prep the homeownership data. Since housing data often isn't available before 2000 (such as this one, too) we use the crude 2007 ownership for the initial population.

```{r filter ownership}
print(housing_es)

# Get ownership rate
ownership <- housing_es %>%
  mutate(obs = as.numeric(obs)) %>% 
  filter(tenure == "OWN", year == 2007) %>%
  mutate(year = 2000) %>%
  select(year, obs)

```

Now we need to find fertility. Importantly, we need to know the order of children, since that affects transition rates. Since Eurostat doesn't have data on childless women, we use data from the Human fertility database instead (https://www.humanfertility.org/Country/Country?cntr=ESP)

```{r load fertility data}
path <- system.file("extdata", "EPC_2026", "ex_year_age_parity", package = "MicSim")
exp_0 <- read.table(path, 
         skip = 4,                    
         col.names = c("year", "age", "E0x", "E1x", "E2x", "E3x","E4x"),
         fill = TRUE,
         strip.white = TRUE,
         stringsAsFactors = FALSE)


fert_0 <- exp_0 %>%
  mutate(
    # Calculate total exposure for each year-age group
    total_exposure = E0x + E1x + E2x + E3x + E4x,
    # Calculate proportions by parity
    prop_0 = E0x / total_exposure,
    prop_1 = E1x / total_exposure,
    prop_2 = E2x / total_exposure,
    prop_3 = E3x / total_exposure,
    prop_4plus = E4x / total_exposure
  ) %>%
   #Pivot to long format for easier use
  pivot_longer(
    cols = starts_with("prop_"),
    names_to = "parity",
    values_to = "proportion"
  ) %>%
  mutate(
    parity = str_remove(parity , "prop_"),
    parity = case_when(
      parity  == "4plus" ~ "4+",
      TRUE ~ parity 
    ) 
  ) %>% 
    select(year, age, parity , proportion)

head(fert_0)
```
Now we have proportions of women for each parity starting in 1991.

With this, we can start building a synthetic population of 5000 individuals.

```{r building population}
N <- 5000

sample_indices <- sample(
  x = 1:nrow(pop_weights),
  size = N,
  prob = pop_weights$weight,
  replace = TRUE
)

synth <- tibble(
  id = 1:N,
  age = pop_weights$age[sample_indices],
  sex = pop_weights$sex[sample_indices]
)

head(synth)

# Sex distribution
print(synth %>% count(sex) %>% mutate(pct = round(n/sum(n)*100, 1)))
```

Assign our initial population random birthdays:

```{r create initial population random birthdays}
#Assume people can get born over the range we have data for
#We create random birthdays but otherwise keep to the age distribution

#assuming we want to start the simulation in 2000
curr_year <- 2000 

synth <- synth %>%
  mutate(
    # Birth year based on current age
    birth_year = curr_year - age,
    # Random day within that year for variation
    day_of_year = sample(1:365, n(), replace = TRUE),
    # Create birthdate
    birthdate = make_date(birth_year, 1, 1) + days(day_of_year - 1),
    # See if we get correct age
    age_check = curr_year - year(birthdate)
  )

print(synth$birthdate[1:10])
print(synth$age[1:10])
print(synth$age_check[1:10])
any(synth$age_check != synth$age)

#drop age variables we don't need anymore

synth <- synth %>% 
  select(- age_check, -birth_year, -day_of_year)
```
And for the women a fertility parity corresponding to their age in the year 2000 again. 

```{r fertility parity}
parity_2000 <- fert_0 %>% 
  filter(year == 2000) %>% 
  pivot_wider(names_from = parity, values_from = proportion) %>%
  select(age, `0`, `1`, `2`, `3`, `4+`)

#Assume age 55 parity for older women

parity_55 <- parity_2000 %>% 
  filter(age == 55) %>%
  select(`0`, `1`, `2`, `3`, `4+`)
max_age <- 100
parity_2000_filled <- parity_2000 %>%
  bind_rows(
    crossing(age = 56:max_age, parity_55)
  ) %>%
  bind_rows(
    tibble(age = 0:11, `0` = 1, `1` = 0, `2` = 0, `3` = 0, `4+` = 0)
  )

#Join

head(synth)

synth_par <- synth %>% 
  filter(sex == "f") %>% 
  left_join(parity_2000_filled, by = "age") %>%
    mutate(across(c(`0`, `1`, `2`, `3`, `4+`), ~replace_na(., 0))) %>%
  rowwise() %>% 
  mutate(
    parity = if_else(age < 12, "0",
                     sample(c("0", "1", "2", "3", "4+"),
                            size = 1,
                            prob = c(`0`, `1`, `2`, `3`, `4+`)))
  ) %>%
  ungroup() %>%
  select(id, age, sex, birthdate, parity)

```

Plot structure:

```{r plots of Age Structure}
# Females by age bin
females <- synth_par %>%
  mutate(age_bin = cut(age, 
                       breaks = seq(0, 105, 5), 
                       labels = paste0(seq(0, 100, 5), "-", seq(4, 104, 5)),
                       right = FALSE)) %>%
  group_by(age_bin) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(sex = "Female", population = count)

# Males by age bin
males <- synth %>%
  filter(sex == "m") %>%
  mutate(age_bin = cut(age, 
                       breaks = seq(0, 105, 5), 
                       labels = paste0(seq(0, 100, 5), "-", seq(4, 104, 5)),
                       right = FALSE)) %>%
  group_by(age_bin) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(sex = "Male", population = -count)

# Combine
df_pyr <- bind_rows(males, females)

# Plot
ggplot(df_pyr, aes(x = population, y = age_bin, fill = sex)) +
  geom_bar(stat = "identity", width = 0.8) +
  geom_vline(xintercept = 0, linewidth = 0.5) +
  scale_x_continuous(labels = function(x) abs(x)) +
  scale_fill_viridis_d(option = "viridis") +
  labs(x = "Population", y = "Age", 
       title = "Synth Spain: Population pyramid 2000") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())

```

We can also plot this:

```{r plot population pyramid}
# Females by age and parity
females_parity <- synth_par %>%
  group_by(age, parity) %>%
  summarise(count = n(), .groups = "drop") %>%
  rename(Alter = age, Parity = parity, Population = count) %>%
  mutate(Geschlecht = "f")

# Males (no parity)
males <- synth %>%
  filter(sex == "m") %>%
  group_by(age) %>%
  summarise(count = n(), .groups = "drop") %>%
  rename(Alter = age, Population = count) %>%
  mutate(Geschlecht = "m", Parity = "Male")

# Combine and negate males
df_pyramid <- bind_rows(females_parity, males) %>%
  mutate(Population = ifelse(Geschlecht == "m", -Population, Population))

# Relative within sex
sex_totals <- df_pyramid %>%
  group_by(Geschlecht) %>%
  summarise(total_sex = sum(abs(Population), na.rm = TRUE), .groups = "drop")

df_pyramid_rel <- df_pyramid %>%
  left_join(sex_totals, by = "Geschlecht") %>%
  mutate(Population_rel = Population / total_sex)

# Plot
ggplot(df_pyramid_rel,
       aes(x = Alter, y = Population_rel, fill = Parity)) +
  geom_bar(stat = "identity", width = 1, position = "stack") +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  coord_flip() +
  scale_y_continuous(labels = function(x) abs(x)) +
  scale_fill_viridis_d(name = "Parity", direction = -1)  +
  labs(x = "Age", y = "Population proportion") +
  theme_minimal(base_size = 13)
```

Lastly, individuals in the population can be either homeowner or not.
For the initial population we don't account for anything more complicated other
than restricting individuals younger than 20 from becoming owners.

```{r homeowner building}
synth <- synth %>%
  mutate(
    homeowner = if_else(age < 20, "r", 
                        if_else(rbinom(n(), size = 1, prob = ownership$obs / 100) == 1, "h", "r"))
  )
```

So, this initial population includes age, fertility and whether they own a home or not. 

Because MicSim takes special input we are going to reformat this initPop into the input needed by MicSim:

```{r building initPop}
synth_full <- synth %>%
  left_join(synth_par %>% select(id, parity), by = "id") %>%
  mutate(parity = replace_na(parity, "0")) #men get assigned 0 as a parity

initpop_epc_example <- synth_full %>%
  mutate(
    birthdate = as.numeric(format(birthdate, "%Y%m%d")),
    initState = str_c(sex, homeowner, parity, sep = "/")
  ) %>%
  select(id, birthdate, initState)

#save the init pop data
#save(initpop_epc_example, file = "initpop_epc_example.rda")
```

For the transitions rates, we also need data. Here we prep this:

We do this for fertility by getting data from the human fertility database (https://www.humanfertility.org/Country/Country?cntr=ESP#age) and add our
calculated parity probabilities.

```{r load data}
path <- system.file("extdata", "EPC_2026", "es_asfr", package = "MicSim")
fert_es <- read.table(path, 
         skip = 4,                    
         col.names = c("year", "age", "asfr"),
         fill = TRUE,
         strip.white = TRUE,
         stringsAsFactors = FALSE)

fert_es$age[fert_es$age %in% "12-"] <- 12
fert_es$age[fert_es$age %in% "55+"] <- 55
```

Now, since fertility transitions also depend on which number of child it is, we
take parity into account again:

```{r parity fertility}
# Parity proportions by age and year (from HFD, 1991-2023)
shares <- fert_0 %>%
  select(year, age, parity, proportion) %>%
  pivot_wider(names_from = parity, values_from = proportion) %>%
  transmute(
    year = as.integer(year),
    age = as.integer(age),
    prop_0 = as.numeric(`0`),
    prop_1 = as.numeric(`1`),
    prop_2 = as.numeric(`2`),
    prop_3 = as.numeric(`3`),
    prop_4plus = as.numeric(`4+`)
  )

# Prepare ASFR data
fert_rate <- fert_es %>%
  mutate(
    age = as.integer(age),
    year = as.integer(year),
    asfr = as.numeric(asfr)
  ) %>%
  filter(age >= 15, age <= 49, year >= 2000) #49 as cutoff age for fertility

# Decompose ASFR by parity, matching year and age
fert_parity <- fert_rate %>%
  left_join(shares, by = c("year", "age")) %>%
  mutate(
    asfr_0  = asfr * prop_0,
    asfr_1  = asfr * prop_1,
    asfr_2  = asfr * prop_2,
    asfr_3  = asfr * prop_3,
    asfr_4plus = asfr * prop_4plus
  ) %>%
  select(year, age, asfr, asfr_0, asfr_1, asfr_2, asfr_3, asfr_4plus)

head(fert_parity)
#save(fert_parity, file = "epc_fertility.rda")
```

We can quickly plot this rate:

```{r plot fertility rate}
plot_dat <- fert_parity %>%
  mutate(
    age = as.numeric(age),
    year = as.numeric(year)
  ) %>%
  filter(
    age >= 15, age <= 49,
    year >= 2000, year <= 2023
  ) %>%
  pivot_longer(
    cols = c(asfr_0, asfr_1, asfr_2, asfr_3, asfr_4plus),
    names_to = "parity",
    values_to = "rate"
  ) %>%
  mutate(
    parity = recode(
      parity,
      "asfr_0"     = "0 ??? 1",
      "asfr_1"     = "1 ??? 2",
      "asfr_2"     = "2 ??? 3",
      "asfr_3"     = "3 ??? 4",
      "asfr_4plus" = "4+ ??? 4+"
    )
  )

ggplot(plot_dat,
       aes(x = age, y = rate, group = year, colour = factor(year))) + 
  geom_line(linewidth = 0.5) +
  facet_wrap(~ fct_relevel(parity, "0 ??? 1", "1 ??? 2", "2 ??? 3", "3 ??? 4", "4+ ??? 4+"), 
             ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(15, 49, 5)) +
  scale_color_viridis_d(name = "Year") +
  labs(x = "Age", y = "Parity-specific fertility rate") +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(size = 12, face = "bold")
  )
```

And mortality from the mortality database (https://mortality.org/Country/Country?cntr=ESP):

```{r mortality}
path <- system.file("extdata", "EPC_2026", "es_m1x1", package = "MicSim")
mort_es <- read.table(path, 
                 skip = 3,                    
                 col.names = c("year", "age", "female", "male", "total"),
                 fill = TRUE,
                 strip.white = TRUE,
                 stringsAsFactors = FALSE)

summary(mort_es)
head(mort_es)

#save(mort_es, file = "epc_mortality.rda")
```

Plot:

```{r plot mortality rate}
# Convert character columns to numeric
mort_plot <- mort_es %>%
  mutate(
    across(c(female, male, total), ~na_if(., ".")),
    female = as.numeric(female),
    male = as.numeric(male),
    total = as.numeric(total),
    age = as.numeric(age)
  ) %>%
  filter(!is.na(age) & !is.na(female) & !is.na(male)) %>% 
  filter(age >= 0 & age <= 99)

# Reshape to long format
mort_plot <- mort_plot %>%
  pivot_longer(cols = c(female, male), names_to = "sex", values_to = "rate")

mort_plot <- mort_plot %>%
  mutate(year_group = floor(year / 20) * 20)

# Plot
ggplot(mort_plot, aes(x = age, y = rate, color = factor(year_group), group = interaction(year, sex))) +
  geom_line() +
  facet_wrap(~ sex, ncol = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  labs(x = "Age", y = "Mortality Rate", color = "Year") +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(size = 12, face = "bold")
  )
```

