---
title: "EPC_2026_Linked_Lives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EPC_2026_Linked_Lives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicSim)
library(tidyverse)
```

Much of this code follows the same setup as our earlier example:

```{r creating variables}
invisible(gc())
rm(list = ls())
set.seed(321342)

maxAge <- 100

# Defining simulation horizon
startDate <- 20000101 # yyyymmdd
endDate   <- 20231231 # yyyymmdd
simHorizon <- c(startDate=startDate, endDate=endDate)

sex <- c("m", "f")
homeowner <- c("r", "h") #simplification!
fert <- c("0","1","2","3", "4+")

stateSpace <- expand.grid(sex=sex, homeowner = homeowner, fert=fert)
absStates <- "dead"
```

## Loading initial population

```{r initial population}
initpop_epc_example <- data(initpop_epc_example)

initpop <- initpop_epc_example %>% 
  rename(ID = id, birthDate = birthdate, initState = initState) %>% 
  as.data.frame()
```

## Defining Transition Rates

```{r load mortality data}
epc_mortality <- data(epc_mortality)

unique(mort_es$age)

mort_es <- mort_es %>%
  filter(age != "110+") %>% 
  filter(year >= 2000) %>%
  mutate(
    age = as.numeric(age),
    female = as.numeric(female),
    male = as.numeric(male)
  ) %>%
  filter(age <= 100)

# Create matrices
mort_male <- mort_es %>%
  select(age, year, male) %>%
  pivot_wider(names_from = year, values_from = male) %>%
  column_to_rownames("age") %>%
  as.matrix()

mort_female <- mort_es %>%
  select(age, year, female) %>%
  pivot_wider(names_from = year, values_from = female) %>%
  column_to_rownames("age") %>%
  as.matrix()

# Rate functions
mortRates_m<- function(age, calTime) {
  rate <- mort_male[as.character(as.integer(age)), as.character(as.integer(calTime[1]))]
  return(rate)
}

mortRates_f <- function(age, calTime) {
  rate <- mort_female[as.character(as.integer(age)), as.character(as.integer(calTime[1]))]
  return(rate)
}

```

Fertility:

```{r load fertility data}
epc_fertility <- data(epc_fertility)

# Filter fertility data
fert_parity <- fert_parity %>%
  complete(
    age = 0:100,
    year = unique(fert_parity$year),
    fill = list(asfr_0 = 0, asfr_1 = 0, asfr_2 = 0, asfr_3 = 0, asfr_4plus = 0)
  ) %>%
  mutate(
    across(starts_with("asfr_"), 
           ~ if_else(age < 15 | age >= 49, 0, .))
  )

# Create matrices
fert_0_1 <- fert_parity %>%
  select(age, year, asfr_0) %>%
  pivot_wider(names_from = year, values_from = asfr_0) %>%
  column_to_rownames("age") %>%
  as.matrix()

fert_1_2 <- fert_parity %>%
  select(age, year, asfr_1) %>%
  pivot_wider(names_from = year, values_from = asfr_1) %>%
  column_to_rownames("age") %>%
  as.matrix()

fert_2_3 <- fert_parity %>%
  select(age, year, asfr_2) %>%
  pivot_wider(names_from = year, values_from = asfr_2) %>%
  column_to_rownames("age") %>%
  as.matrix()

fert_3_4 <- fert_parity %>%
  select(age, year, asfr_3) %>%
  pivot_wider(names_from = year, values_from = asfr_3) %>%
  column_to_rownames("age") %>%
  as.matrix()

# Rate functions
fertRates_0_1 <- function(age, calTime) {
  rate <- fert_0_1[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

fertRates_1_2 <- function(age, calTime) {
  rate <- fert_1_2[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

fertRates_2_3 <- function(age, calTime) {
  rate <- fert_2_3[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

fertRates_3_4 <- function(age, calTime) {
  rate <- fert_3_4[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

```

For the homeownership rate in a scenario with intergenerational transmission, the function requires data for the years when the mother would be transitioning. Since these data are not present in our example homeownership matrix, we backfilled using the 2005 data that exists.

```{r homeownership rate}
epc_ho_matrix <- data(epc_ho_matrix)

ownhome <- function(age, calTime) {
  age_int <- as.integer(age)
  year_idx <- as.character(min(as.integer(calTime[1]), 2020))
  
  rate <- ho_matrix[as.character(age_int), year_idx]
  rate[age_int < 20 | age_int > 70] <- 0
  return(rate)
}

```

We plot this rate:

```{r rate plot}
# Average across all years by age
ho_avg <- rowMeans(ho_matrix, na.rm = TRUE)
age <- as.numeric(rownames(ho_matrix))


plot(age, ho_avg, 
     type = "l", 
     lwd = 2, 
     col = "steelblue",
     xlab = "Age", 
     ylab = "Homeownership rate",
     main = "Mean homeownership rate by age")
```

Again we construct the matrices `micSim` needs:

```{r building matrices}
fertTrMatrix <- cbind(c("f/0->f/1",
                        "f/1->f/2",
                        "f/2->f/3", 
                        "f/3->f/4+"),
                      c("fertRates_0_1", 
                        "fertRates_1_2", 
                        "fertRates_2_3", 
                        "fertRates_3_4" ))

homatrix <- cbind("r->h", "ownhome")

allTransitions <- rbind(fertTrMatrix, homatrix)

absTransitions <- cbind(c("f/dead","m/dead"),
                      c("mortRates_f", "mortRates_m"))

```

The transition matrix:

```{r transition matrix}
transitionMatrix <- buildTransitionMatrix(allTransitions=allTransitions,
                                          absTransitions=absTransitions,
                                          stateSpace=stateSpace)

head(transitionMatrix)
```

What we need for fertility:

```{r fertility matrix}
fertTr <- fertTrMatrix[,1]
print(fertTr)
```

```{r childhood states}
varInitStates <- rbind(c("m","r", "0"),c("f", "r", "0"))
initStatesProb <- c(0.515,0.485)
```

## Adding in intergenerational transmission

If you want to link generations together, this is all that is needed:

The $alpha$ is set to 0 in this case and our $beta$ was estimated using data from @beduk2025a.
In `geninfo` you supply the transition rate that is considered for intergenerational transmission, which ID the individual is linked to, as well as the $alpha$ and $beta$ which are used to transform the original rate of, in this case, the mother.

```{r generation matrix}
geninfo <- c("ownhome", "motherID", 0, 1.45) 

genMatrix <- build_genMatrix(geninfo)

print(genMatrix)
```

Now we run the microsimulation, but again, we reduce the amount of individuals in our population. Also, this time we use the new function `micSimLink`.

```{r simulation}
set.seed(321342)
#It takes about 2 minutes to complete the simulation!
#You can increase the population up to 5000 for better plotting
poplink <- micSimLink(initPop=initpop[1:1000,],
                      transitionMatrix=transitionMatrix,
                      absStates=absStates,
                      varInitStates=varInitStates,
                      initStatesProb=initStatesProb,
                      maxAge=maxAge, simHorizon=simHorizon,
                      fertTr=fertTr, genMatrix = genMatrix)
```

Let's look at our output again:

```{r simulation long format}
#pop_2 <- poplink %>% filter(!is.na(motherID))

pop_long <- convertToLongFormat(poplink, migr = FALSE)
head(pop_long, 3)

```

```{r plot homeownership}
pop_plot_h <- pop_long %>%  
  mutate(
    agestart = getAgeInDays(Tstart, birthDate) / 365.25,
    agestop = getAgeInDays(Tstop, birthDate) / 365.25,
    year = as.numeric(substr(Tstop, 1, 4))
  )  %>%
  filter(OD == "r->h") %>% 
  group_by(ID) %>% 
  slice_max(agestop, na_rm = TRUE) %>% 
  ungroup()

ggplot(pop_plot_h, aes(x = agestop)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  labs(x = "Age at homeownership transition", 
       y = "Density",
       title = "Age distribution of transitions to homeownership") +
  theme_minimal(base_size = 13)
```

In an even more complex example it would also be possible to match mothers and fathers together. The syntax for this can also be found in the other examples in the `micSim` package.
