---
title: "EPC_2026_Microsimulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EPC_2026_Microsimulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicSim)
library(tidyverse)
library(glue)
```

In this vignette we will run a simple microsimulation. 
You can also check out the examples of the `MicSim()` and `MicSimlink()` functions for more.

## Definition of the Simulation Frame

These are the basics that all microsimulation needs: 
- max Age: the maximum age that people can reach in a simulation
- Simulation horizon: from when until when the simulation runs
- State Space: the collection of attributes that individuals can take in the simulation

```{r creating variables}
invisible(gc())
rm(list = ls())
set.seed(321342)

maxAge <- 100

# Defining simulation horizon
startDate <- 20000101 # yyyymmdd
endDate   <- 20231231 # yyyymmdd
simHorizon <- c(startDate=startDate, endDate=endDate)

sex <- c("m", "f")
homeowner <- c("h", "r") #simplification!
fert <- c("0","1","2","3", "4+")

stateSpace <- expand.grid(sex=sex, homeowner = homeowner, fert=fert)
absStates <- "dead"
```

## Reading in the Initial Population
Loading in data and making sure that everything is named correctly.
When using `MicSim` the `initpop` **has** to be named exactly like you can see below, including the capitalization.

```{r loading initial population}
#data(initpop_epc_example)
load("EPC_data/initpop_epc_example.rda")

initpop <- initpop_epc_example %>% 
  rename(ID = id, birthDate = birthdate, initState = initState) %>% 
  as.data.frame()
```

## Defining Transition Rates

In this step we define our transition rate functions using data prepared before.
The data preparation is documented in the corresponding vignette "EPC_2026_Data_Preparation".

We start with mortality:
First we clean up the data we use for this example a little. 
These are just a matrix containing the age-period rate of mortality for men and women separately.

These matrices are used then by the corresponding transition rate **function**.
`MicSim` needs those functions to tell it where to find the transition rate input.
In this example, the function looks up the correct value in the corresponding matrix.
For your research you might want to define transition rates differently, which is fine.

Importantly, a transition rate always has to be given for at least the whole age range covered in `MicSim` [0,`maxAge`].
Calendar time, `calTime`, has to be given as an argument, but doesn't necessarily need to be reflected in the function (in most cases you probably want it to be).

```{r transition rates}
#data(epc_mortality)
epc_mortality <- load("EPC_data/epc_mortality.rda")

unique(mort_es$age)

mort_es <- mort_es %>%
  filter(age != "110+") %>% 
  filter(year >= 2000) %>%
  mutate(
    age = as.numeric(age),
    female = as.numeric(female),
    male = as.numeric(male)
  ) %>%
  filter(age <= 100)

# Create matrices
mort_male <- mort_es %>%
  select(age, year, male) %>%
  pivot_wider(names_from = year, values_from = male) %>%
  column_to_rownames("age") %>%
  as.matrix()

mort_female <- mort_es %>%
  select(age, year, female) %>%
  pivot_wider(names_from = year, values_from = female) %>%
  column_to_rownames("age") %>%
  as.matrix()

# Rate functions
mortRates_m<- function(age, calTime) {
  rate <- mort_male[as.character(as.integer(age)), as.character(as.integer(calTime[1]))]
  return(rate)
}

mortRates_f <- function(age, calTime) {
  rate <- mort_female[as.character(as.integer(age)), as.character(as.integer(calTime[1]))]
  return(rate)
}
```

We do the same thing we did for fertility also for mortality:
- Loading data in
- Creating matrices of data for transition rates
- Defining transition rate functions

```{r loading fertility data}
#data(epc_fertility)
epc_fertility <- load("EPC_data/epc_fertility.rda")

# Filter fertility data
fert_parity <- fert_parity %>%
  complete(
    age = 0:100,
    year = unique(fert_parity$year),
    fill = list(asfr_0 = 0, asfr_1 = 0, asfr_2 = 0, asfr_3 = 0, asfr_4plus = 0)
  ) %>%
  mutate(
    across(starts_with("asfr_"), 
           ~ if_else(age < 15 | age >= 49, 0, .))
  )

# Create matrices
fert_0_1 <- fert_parity %>%
  select(age, year, asfr_0) %>%
  pivot_wider(names_from = year, values_from = asfr_0) %>%
  column_to_rownames("age") %>%
  as.matrix()

fert_1_2 <- fert_parity %>%
  select(age, year, asfr_1) %>%
  pivot_wider(names_from = year, values_from = asfr_1) %>%
  column_to_rownames("age") %>%
  as.matrix()

fert_2_3 <- fert_parity %>%
  select(age, year, asfr_2) %>%
  pivot_wider(names_from = year, values_from = asfr_2) %>%
  column_to_rownames("age") %>%
  as.matrix()

fert_3_4 <- fert_parity %>%
  select(age, year, asfr_3) %>%
  pivot_wider(names_from = year, values_from = asfr_3) %>%
  column_to_rownames("age") %>%
  as.matrix()

# Rate functions
fertRates_0_1 <- function(age, calTime) {
  rate <- fert_0_1[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

fertRates_1_2 <- function(age, calTime) {
  rate <- fert_1_2[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

fertRates_2_3 <- function(age, calTime) {
  rate <- fert_2_3[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

fertRates_3_4 <- function(age, calTime) {
  rate <- fert_3_4[as.character(as.integer(age)),
                   as.character(as.integer(calTime[1]))]
  return(rate)
}

```

And finally a homeownership rate. This one is the only one that comes from data that are not publicly available.
Again, we follow a similar process, except that we already have the correct matrix format.

```{r create homeownership function}
#data(epc_ho_matrix)
epc_ho_matrix <- load("EPC_data/epc_ho_matrix.rda")

ownhome <- function(age, calTime) {
  age_int <- as.integer(age)
  year_idx <- as.character(min(as.integer(calTime[1]), 2020))
  
  rate <- ho_matrix[as.character(age_int), year_idx]
  rate[age_int < 20 | age_int > 70] <- 0
  return(rate)
}

```

We plot this rate (we plot the mean for easier visualization):

```{r plot homeownership}
# Average across all years by age
ho_avg <- rowMeans(ho_matrix, na.rm = TRUE)
age <- as.numeric(rownames(ho_matrix))

plot(age, ho_avg, 
     type = "l", 
     lwd = 2, 
     col = "steelblue",
     xlab = "Age", 
     ylab = "Homeownership rate",
     main = "Mean homeownership rate by age")
```

Now we need to construct a few other matrices that `micSim` needs:

These all transport the information of which transition rate function corresponds to which transition.
We differentiate between non-absorbing and absorbin states (`dead` is the absorbing state considered here).

```{r matrices}
fertTrMatrix <- cbind(c("f/0->f/1",
                        "f/1->f/2",
                        "f/2->f/3", 
                        "f/3->f/4+"),
                      c("fertRates_0_1", 
                        "fertRates_1_2", 
                        "fertRates_2_3", 
                        "fertRates_3_4" ))

homatrix <- cbind("r->h", "ownhome")

allTransitions <- rbind(fertTrMatrix, homatrix)

absTransitions <- cbind(c("f/dead","m/dead"),
                      c("mortRates_f", "mortRates_m"))

```

The `micSim` package has a function (`buildTransitionMatrix`) to bind all the individual matrices together into the correct input `MicSim` needs:
Always double check that this matrix is actually correct!

```{r transition matrix}
transitionMatrix <- buildTransitionMatrix(allTransitions=allTransitions,
                                          absTransitions=absTransitions,
                                          stateSpace=stateSpace)

head(transitionMatrix)
```

Since we provide fertility rates, we also need to tell `micSim` which transitions control fertility:

```{r fertility matrix}
fertTr <- fertTrMatrix[,1]
print(fertTr)
```

We provided fertility, so `micSim`will add newborns. 
This means, we need to tell `micSim` which states it should assign to children.
For this simulation we only want newborns to differ between being "male" and "female".
So we assign these states with the probabilities below:

```{r childhood states}
varInitStates <- rbind(c("m","r", "0"),c("f", "r", "0"))
initStatesProb <- c(0.515,0.485)
```

Since this microsimulation does not use any links, we can use the standard `micsim` function. We have restricted the initpop to the first 500, to make computation quicker for this workshop.
In general, this is a very short simulation, depending on you research question you may want a longer simulation horizon, but this will again increase computational resources. We run this microsimulation with the full population.

```{r simulation}
set.seed(321342)

pop <- micSim(initPop=initpop[1:500,],
              transitionMatrix=transitionMatrix, absStates=absStates,
              varInitStates=varInitStates, initStatesProb = initStatesProb, maxAge=maxAge, simHorizon=simHorizon, fertTr=fertTr)
```

Now that we have an output, we can see what it look like 
(we remove censored people here):

```{r table}
pop %>%
  filter(is.na(motherID)) %>%
  filter(!if_any(-motherID, is.na)) %>%
  head()
```

As you can see, `MicSim` has a very specific output. 
We can make this more readable by converting it to a long format.
This makes it easier to see how long individuals spent in each state:

```{r simulation long format}
long <- convertToLongFormat(pop, migr = FALSE)
head(long)
```

If we want to do some basic plots, we should make age more readable:

```{r df for plotting}
pop_plot <- long %>%  mutate(
    agestart = getAgeInDays(Tstart, birthDate) / 365.25,
    agestop = getAgeInDays(Tstop, birthDate) / 365.25,
  )
  
```

Since we only had very few individuals in our simulation over a short time, 
there won't be that many transitions captured. Still, we can plot fertility and
mortality:

```{r fertility plot, warning=FALSE}

plot_fertility <- pop_plot %>%
  filter(grepl("^[0-9]->", OD)) %>%
  mutate(
    od_orig = as.numeric(substr(OD, 1, 1)),
    od_dest = as.numeric(substr(OD, 4, 4))
  ) %>%
  filter(fert == od_orig)

ggplot(plot_fertility, aes(agestop, fill = factor(od_dest))) +  
  geom_density(alpha = 0.4) +  
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.8) +  
  labs(title = "Age at fertility transitions", 
       x = "Age at transition", 
       y = "Density",
       fill = "Transition to") +  
  theme_minimal(base_size = 12) +  
  theme(legend.position = "right",  
        panel.grid.major.x = element_blank())
```

```{r mortality plot}

pop_plot_dead <- pop_plot %>%
  filter(OD == "dead")

ggplot(pop_plot_dead, aes(agestop, fill = sex)) +  
  geom_density(alpha = 0.6) +  
  scale_fill_viridis_d(option = "plasma", begin = 0.25, end = 0.85) +
  labs(x = "Age at death", y = "Density") +  
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

```


Feel free to experiment yourself!
